### 2019-02-13 manuscript/methods.tex
started editing Methods

Chromosome number data were obtained for all Solanaceae taxa in the the Chromosome Counts Database \citep[CCDB;][]{rice_2015}, and the ca.\ 14,000 records were cleaned semi-automatically using the CCDBcurator R package \citep{rivero_2019}. 
This large dataset includes the compilation of Solanaceae ploidy states from \citep{robertson_2011}.
Species were coded as either diploid (D) or polyploid (P).
For the majority of species, ploidy was assigned according to information from the original publications and the Kew Royal Botanic Gardens C-value DNA resource \citep{bennett_2005}.
For taxa without ploidy information but with information about chromosome number, we assigned ploidy based on the multiplicity of chromosomes within the genus.
For example, \textit{Solanum betaceum} did not include information about ploidy level but it has 24 chromosomes, so because $x=12$ is the base chromosome number of the \textit{Solanum} genus \citep{olmstead_2007}, we assigned \textit{S.~betaceum} as diploid. 
Species with more than one ploidy level were assigned the smallest and most frequent ploidy level recorded.  We also cross-verified using the breeding system information and found that there were no self-incompatible species possibly listed as polyploid, or listed as polyploid and diploid which increase our confidence in the assignations% E: Does this apply to the diploid-polyploid levels? R- Yes there are no self-incompatible polyploids at all.  If so, we need to explain how it is also reconciled against the SI/SC value.  Or is this sentence just about multiple levels of polyploidy, which wouldn't need to be mentioned here? R:  Solved
Breeding system was scored as self-incompatible (I) or self-compatible (C) based on results hand-curated from the literature (as in \citealt{igic_2006, goldberg_2010, robertson_2011, goldberg_2012}).
Most species could unambiguously be coded as either SI or SC \citep{raduski_2012}.
Following previous work, we coded as I any species with functional I systems, even if C or dioecy was also reported.
Dioecious species without functional I were coded as C.

To those existing data sets, we added some additional records for chromosome number and breeding system.
The Supplementary Information contains citations for the numerous original sources for all of the data. % todo: add number of refs when we have it
Synonymy followed Solanaceae Source \citep{solsource}.
Hybrids and cultivars were excluded because ploidy and breeding system are likely to be altered by domestication.
As in \citet{robertson_2011}, we examined closely the few species for which the merged ploidy and breeding system data indicated the presence of I polyploids.
Although SI populations frequently contain some C individuals and diploid populations frequently contain some polyploid individuals, in no case did we find individuals that were both SI and polyploid.
Because of this empirical observation and the functional explanation for whole genome duplication disabling gametophytic self-incompatibility \citep[reviewed in][]{ramsey_1998,stone_2002}, we consider only three observed character states: self-incompatible diploid (ID), self-compatible diploid (CD), and polyploid (P) which are always self-compatible.

% E: We should probably include some more details about how many polymorphic species and how the particularly tricky ones were dealt with.

Matching our character state data to the largest time-calibrated phylogeny of Solanaceae \citep{sarkinen_2013} yielded 595 species with ploidy and/or breeding system information on the tree.
Binary or three-state classification of ploidy and breeding system for the 595 taxa is summarized in \cref{table:stateclassifications}. % E: I think you're still updating the state coding? R=no since models are not comparable there is no need. 
We retained all of these species in each of the analyses below, because pruning away tips lacking breeding system in the ploidy-only analyses (and vice versa) would discard data that could inform the diversification models. A total of 405 taxa without any information about breeding system nor polyploidy were pruned. Tips without trait data are non informative for  diversification parameters linked to trait change and can significantly slow-down the convergence of inferences in trees with already hundreds of tips.% E: Though someone might ask why tips with no state data were pruned.  Analyses took long enough on even this-size tree. R: I think this point needs to be further investigated in another paper, so I just sort-of justify our choice

\subsection{Models for Diversification and Polyploidy}

% E: There are so many models, I think we need to try hard to help the reader keep them straight.
% I would avoid naming any states with numbers, except maybe within the stateclassification table and its caption.
% I also still find bisse/hisse/muhisse terminology confusing, I think because it both describes a general class of model and applies to a specific analysis here, and because bisse and hisse are run separately on ploidy breeding system.
% How about a model naming scheme like this?
%   D/P polyploidy model
%   D/P-A/B polyploidy hidden state model
%   I/C breeding system model
%   I/C-A/B breeding system and hidden state model
%   ID/CD/CP ploidy and breeding system model
%   ID/CD/CP-A/B ploidy, breeding system and hidden state model
% Or something else more concise and descriptive?

We first defined a binary state speciation and extinction model (BiSSE, \citet{maddison_2007})  for polyploidy evolution where taxa was classified as diploid (D=0) and polyploid (P=1) \cref{table:stateclassifications}. 
We refer to this model as the D/P polyploidy model. 
In a Bayesian framework, we obtained posterior probability distributions of speciation rates ($\lambda_D$, $\lambda_P$), extinction rates ($\mu_D$, $\mu_P$) , net diversification rates ($r_D=\lambda_D-\mu_D, r_P=\lambda_P-\mu_P$), and  relative extinction rates ($\nu_D=\frac{\mu_D}{\lambda_D}, \nu_D=\frac{\mu_D}{\lambda_D}$) just as previously explored in \citet{mayrose_2011} and \citet{mayrose_2015}. However, our analyses differ from previous models since we consider not only a polyploidization rate $\rho$ and but also a diploidization parameter rate $\delta$.

As a second step, we fitted a hidden state speciation and extinction model (HiSSE, \citet{beaulieu_2016}) to evaluate whether the differences in diversification rates were found due to a hidden trait associated to polyploidy.
This model is referred to as the D/P-A/B polyploidy and hidden state model.
As discussed by \citet{beaulieu_2016} BiSSE-like models suffer from a large type I error since they fail to account as part of the null hypothesis heterogeneous diversification rate changes  that do not directly depend on the trait of interest.
By expanding the D/P model by using a hidden trait with states A and B,  the polyploidy and hidden state model addressed the signal of diversification due to polyploidy while parsing out the heterogeneity of diversification rates due to unobserved traits.
The D/P-A/B polyploidy and hidden trait is composed by four states: diploid and polyploid subdivided by a binary hidden trait with states A and B for which we estimated the posterior probability distributions of speciation rates ($\lambda_{D_A},\lambda_{D_B}, \lambda_{P_A},\lambda_{P_B}$), extinction rates ($\mu_{D_A},\mu_{D_B}, \mu_{P_A},\mu_{P_B}$),  net diversification rates ($r_{D_A},r_{D_B},r_{P_A},r_{P_B}$), and relative extinction rates($\nu_{D_A},\nu_{D_B}, \nu_{PA},\nu_{P_B}$). In this model polyploidization rate $\rho$, diploidization rate $\delta$ are also considered, while changes between every hidden state are symmetrical with the same rate $\alpha$.

\subsection{Models for Diversification and Breeding System}
In order to analyze the effects of breeding system in the diversification process, two models the I/C (I= self-incompatible, C= self-compatible) and the I/C-A/B were performed. The I/C breeding system model (BiSSE) was done with the goal of investigating if the effect of self-incompatibility in diversification similar to the analysis performed by \citet{goldberg_2012}. The I/C-A/B  breeding system with hidden trait model parses out the effect of breeding system in diversification by incorporating the possibility of heterogenous changes in the diversification rates that can be attributed to a hidden state. A hidden state model has not previously been fitted to Solanaceae data. For these two models, we assumed that self-compatibility is irreversible as discussed by \citet{igic_2013}, so we defined the transition rate $q_{IC}$ as the parameter for changes from self-incompatible to self-compatible state. This is an important assumption because ancestral state reconstruction in the models might differ when polyploidy and breeding system are analyzed as independent  in the Solanaceae tree. This irreversibility assumption prevents self-compatible state from evolving into self-incompatible state and as a result, the ancestral state of all taxa in this model has probability one of being self-incompatible.   % E: explain irreversibility assumption; fix ancestry instead? - R: Fix ancestry doesn't prevent you from going back to SI from SC but irreversibility conditions  the ancestral state to SI by default. I thought the latter is what we wanted based on Igic et al 2013 correct?
%The input data for these models was coded as self-compatible (C=0) or self-incompatible (I=1) \cref{table:stateclassifications}.

\subsection{Models for Polyploidy, Breeding System, and Diversification}
As a third approach, we analyzed the roles of both polyploidy and breeding system in the process of diversification using a multivariate speciation and extinction model (MuSSE, \citet{fitzjohn_2012}).  We named our MuSSE the ID/P/CD breeding system and polyploidy model, where the three states are self-incompatible diploids (ID), polyploids (always self-compatible, P=1), and self-compatible diploids (CD). The model is defined via ten parameters; six defining diversification ($\lambda_{CD}, \mu_{CD},\lambda_{P},\mu_{P}, \lambda_{ID},\mu_{ID}$) and  four parameters defining key trait changes: polyploidization of self-compatible diploids $\rho_{C}$, diploidization $\delta$, polyploidization of self-incompatible diploids $\rho_{I}$, and self-incompatible to self-compatible rate $q_{IC}$.

>>>>

Chromosome number data were obtained for all Solanaceae taxa in the the Chromosome Counts Database \citep[CCDB;][]{rice_2015}, and the ca.~14,000 records were cleaned semi-automatically using the CCDBcurator R package \citep{rivero_2019}. 
This large dataset includes the compilation of Solanaceae ploidy states from \citet{robertson_2011}.
Species were coded as either diploid (D) or polyploid (P).
For the majority of species, ploidy was assigned according to information from the original publications and the Kew Royal Botanic Gardens C-value DNA resource \citep{bennett_2005}.
For taxa without ploidy information but with information about chromosome number, we assigned ploidy based on the multiplicity of chromosomes within the genus.
For example, \textit{Solanum betaceum} did not include information about ploidy level but it has 24 chromosomes, so because $x=12$ is the base chromosome number of the \textit{Solanum} genus \citep{olmstead_2007}, we assigned \textit{S.~betaceum} as diploid. 
Species with more than one ploidy level were assigned the smallest and most frequent ploidy level recorded.
% E: There was more here about cross-verifying that no P species were SI.  I removed it because there were conflicts before, but we resolved them.  This is addressed in the next paragraph.
%
Breeding system was scored as self-incompatible (I) or self-compatible (C) based on results hand-curated from the literature \citep[as compiled in][]{igic_2006, goldberg_2010, robertson_2011, goldberg_2012}.
Most species could unambiguously be coded as either I or C \citep{raduski_2012}.
Following previous work, we coded as I any species with functional I systems, even if C or dioecy was also reported.
Dioecious species without functional I were coded as C.

% E: I am trying to stick with I/C when talking about the state values, but using SI/SC when talking about the meaning of the trait (because this what people use normally).

To those existing data sets, we added some additional records for chromosome number and breeding system.
The Supplementary Information contains citations for the numerous original sources for all of the data. % todo: add number of refs when we have it
Synonymy followed Solanaceae Source \citep{solsource}.
Hybrids and cultivars were excluded because ploidy and breeding system are likely to be altered by domestication.
As in \citet{robertson_2011}, we examined closely the few species for which the merged ploidy and breeding system data indicated the presence of self-incompatible polyploids.
Although SI populations frequently contain some SC individuals and diploid populations frequently contain some polyploid individuals, in no case did we find individuals that were both SI and polyploid.
Because of this empirical observation and the functional explanation for whole genome duplication disabling gametophytic self-incompatibility \citep[reviewed in][]{ramsey_1998,stone_2002}, we consider only three observed character states: self-incompatible diploids (ID), self-compatible diploids (CD), and polyploids which are always self-compatible (CP).

% E: Should we include more details about how many polymorphic species and how the particularly tricky ones were dealt with?

Matching our character state data to the largest time-calibrated phylogeny of Solanaceae \citep{sarkinen_2013} yielded 595 species with ploidy and/or breeding system information on the tree.
Binary or three-state classification of ploidy and breeding system for the 595 taxa is summarized in \cref{figure:stateclassifications}.
We retained all of these species in each of the analyses below because pruning away tips lacking breeding system in the ploidy-only analyses (and vice versa) would discard data that could inform the diversification models.
A total of 405 taxa without any information about breeding system or polyploidy were excluded.
Tips without trait data are much less informative for diversification parameters linked to trait values, and including this many more species would have prohibitively slowed our analyses, expecially for the most complex models.

\subsection{Models for Ploidy and Diversification}

% D/P polyploidy model
% D/P-A/B polyploidy hidden state model
% I/C breeding system model
% I/C-A/B breeding system and hidden state model
% ID/CD/CP ploidy and breeding system model
% ID/CD/CP-A/B ploidy, breeding system and hidden state model

To investigate the association between ploidy level and diversification, we first defined a binary state speciation and extinction model (BiSSE, \citealt{maddison_2007}) in which taxa were classified as diploid (D) or polyploid (P) (\cref{figure:stateclassifications}).
We call this the $D/P$ ploidy model. 
In a Bayesian framework, we obtained posterior probability distributions of speciation rates ($\lambda_D$, $\lambda_P$), extinction rates ($\mu_D$, $\mu_P$), net diversification rates ($r_D=\lambda_D-\mu_D$, $r_P=\lambda_P-\mu_P$), and relative extinction rates ($\nu_D=\mu_D / \lambda_D$, $\nu_D=\mu_D / \lambda_D$) associated with each state.
This analysis explores the same question as \citet{mayrose_2011, mayrose_2015}, but our analyses differ because we include not only polyploidization (parameter $\rho$, the transition rate from $D$ to $P$), but also diploidization (parameter $\delta$, the transition rate from $P$ to $D$). % E: presumably the motivation will be explained in the Intro

To evaluate whether diversification rate differences found by the previous model could be explained by something other than ploidy, we also added a hidden state (HiSSE model; \citet{beaulieu_2016}).
We call this the $D/P--A/B$ ploidy and hidden state model.

% As a second step, we fit a hidden state speciation and extinction model (HiSSE, \citet{beaulieu_2016}) to evaluate whether the differences in diversification rates were found due to a hidden trait associated to polyploidy.
% This model is referred to as the D/P-A/B polyploidy and hidden state model.
As discussed by \citet{rabosky_2015, beaulieu_2016}, BiSSE-like models can suffer from a large false discovery rate because they fail to account for diversification rate changes that do not directly depend on the trait of interest.
By expanding the $D/P$ model to include a hidden trait with states $A$ and $B$, the ploidy and hidden state model addressed the signal of diversification due to polyploidy while parsing out the heterogeneity of diversification rates due a possible unobserved trait.
The $D/P--A/B$ model is composed by four states: diploid and polyploid subdivided by a binary hidden trait with states A and B for which we estimated the posterior probability distributions of speciation rates ($\lambda_{D_A},\lambda_{D_B}, \lambda_{P_A},\lambda_{P_B}$), extinction rates ($\mu_{D_A},\mu_{D_B}, \mu_{P_A},\mu_{P_B}$),  net diversification rates ($r_{D_A},r_{D_B},r_{P_A},r_{P_B}$), and relative extinction rates($\nu_{D_A},\nu_{D_B}, \nu_{PA},\nu_{P_B}$). In this model polyploidization rate $\rho$, diploidization rate $\delta$ are also considered, while changes between every hidden state are symmetrical with the same rate $\alpha$.

\subsection{Models Breeding System and Diversification}

In order to analyze the effects of breeding system in the diversification process, two models the I/C (I= self-incompatible, C= self-compatible) and the I/C-A/B were performed. The I/C breeding system model (BiSSE) was done with the goal of investigating if the effect of self-incompatibility in diversification similar to the analysis performed by \citet{goldberg_2012}. The I/C-A/B  breeding system with hidden trait model parses out the effect of breeding system in diversification by incorporating the possibility of heterogenous changes in the diversification rates that can be attributed to a hidden state. A hidden state model has not previously been fitted to Solanaceae data. For these two models, we assumed that self-compatibility is irreversible as discussed by \citet{igic_2013}, so we defined the transition rate $q_{IC}$ as the parameter for changes from self-incompatible to self-compatible state. This is an important assumption because ancestral state reconstruction in the models might differ when polyploidy and breeding system are analyzed as independent  in the Solanaceae tree. This irreversibility assumption prevents self-compatible state from evolving into self-incompatible state and as a result, the ancestral state of all taxa in this model has probability one of being self-incompatible.   % E: explain irreversibility assumption; fix ancestry instead? - R: Fix ancestry doesn't prevent you from going back to SI from SC but irreversibility conditions  the ancestral state to SI by default. I thought the latter is what we wanted based on Igic et al 2013 correct?
%The input data for these models was coded as self-compatible (C=0) or self-incompatible (I=1) \cref{table:stateclassifications}.

\subsection{Models for Ploidy, Breeding System, and Diversification}

As a third approach, we analyzed the roles of both polyploidy and breeding system in the process of diversification using a multivariate speciation and extinction model (MuSSE, \citet{fitzjohn_2012}).  We named our MuSSE the ID/P/CD breeding system and polyploidy model, where the three states are self-incompatible diploids (ID), polyploids (always self-compatible, P=1), and self-compatible diploids (CD). The model is defined via ten parameters; six defining diversification ($\lambda_{CD}, \mu_{CD},\lambda_{P},\mu_{P}, \lambda_{ID},\mu_{ID}$) and  four parameters defining key trait changes: polyploidization of self-compatible diploids $\rho_{C}$, diploidization $\delta$, polyploidization of self-incompatible diploids $\rho_{I}$, and self-incompatible to self-compatible rate $q_{IC}$.
